### 2020-12-02 paper/2020-10-revision-background.tex
First go at shorten

This section provides a conceptual framework for understanding how birth spacing changes
and guiding the subsequent analysis.
In non-Western countries, lower fertility is associated with longer birth spacing, but 
outside a ``natural fertility'' regime it is not clear that lower fertility necessarily 
leads to longer spacing \citet{Newman1988,Rutstein2011,Casterline2016}.
Thus, I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I argue that female education and area of residence are the most important factors 
to consider in the empirical analysis and discuss how they tie into the three theories.

The dramatic improvements in economic conditions in India is best illustrated by the 
doubling of real wages between 1987 and 2011 \citep{Klasen2015}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as a good becomes more expensive relative to other
goods, people substitute away from the now more expensive good and towards other goods.
With higher wages, the substitute effect predicts that people work more and spend less 
time on other time-intensive non-wage earning activities, such as children or leisure.
The income effect captures that even if we ignore the substitution effect and hold hours 
worked constant higher wages increase the available income. 
This positive income effect leads people to spend more time on time-intensive activities 
such as children and leisure and less time working.

Empirically, higher female wages generally lowers fertility, whereas higher male 
wages increases fertility \citet{Hotz1997,schultz97}.
This pattern is consistent with women spending substantially more time on child care 
than men, leading to the substitution effect dominating for women while the income
effect dominates for men.

In addition to lowering fertility, higher female wages will tend to shorten spacing between 
children if having children requires the mother to reduce her economic activities.
Specifically, shortening birth spacing allows parents to take advantage of economies of 
scale in childrearing---for example, looking after two children simultaneously requires 
less than double the time needed to look after one child \citep{Vijverberg1982,Hotz1997}.
The reduction in the length of birth intervals from higher female wages will be more 
pronounced the more women participate in the labor force rather than working at home as 
economic activities at home are more compatible with child rearing.

Parents care not only about how many children they have but also about the outcomes
for the children, and, everything else equal, more children means less resources to invest 
per child.
Hence, with declining fertility we should see better child outcomes such as lower 
mortality and more schooling.
However, with increasing income, it is, in principle, possible to have both more
children and better outcomes for them.
Increasing returns to education might make the parents reduce fertility and invest 
more in each of the children's education \citep{Rosenzweig1982a}.

Longer spacing between births is generally expected to have positive effects on 
child outcomes and if fertility falls because of increased desired for better child
outcomes we would, therefore, also expect spacing to increase.
The clearest example of birth spacing affecting child outcomes is for health,
where very short spacing---approximately 24 months or less---leads to worse health 
and mortality outcomes for children 
\citep{Whitworth2002,Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.
More speculative is whether the increased individual parental attention per child that 
should arise from longer spacing results in better human capital outcome for children 
\citep{Zajonc1975,Zajonc1976,Razin1980}.
The evidence on spacing's effect on child quality measures such as IQ and education is 
mixed for developed countries and nonexisting for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

Empirically, son preference has only a minor effect on overall fertility, but does impact 
the sex composition within families through differential stopping behavior 
\citet{repetto72,leung94,clark00,Basu2010,Barcellos2014}. 
Furthermore, as mentioned above, stronger son preference leads to shorter spacing and the 
associated worse health outcomes for daughters when there are no sons in the household 
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.

@@

Holding son preference constant, lower desired fertility increases the use of sex 
selection, suggesting that the first order effect of higher female wages is to increase 
sex selection because of the fertility effect discussed above 
\citep{Portner2015b,Jayachandran2017}.  
Furthermore, the higher the parity the more likely the use of sex selection if the 
desired number of sons has not been reached.
[cost of sex selection?]

% Factors/predictions
Female education and area of residence are the two principal explanatory variables I use
in the empirical analyses.
The main reason is that female education and area of residence are the closest proxies in 
the data for wages and access to and cost of sex selection, none of which are available.
In addition, both higher female education and urbanization are associated with lower 
fertility and increased use of sex selection in past research
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.

Finally, India has seen tremendous increases in both education and urbanization.
The proportion of the population living in urban areas almost doubled from 18\%
in 1960 to 35\% in 2019 \citep{United-Nations2019}.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 20 years or older, whether married or not, in the 
four rounds of the NFHS.
Female education has increased substantially in both rural and urban areas,
but is still considerably higher in urban than in rural areas.
In rural areas, women with eight or more years of education have gone from almost zero for 
the 1930 cohort to more than 60\% for the 1990s cohorts, with about half in the 8--11 
group and the other half in the 12 plus group.
In urban areas, most of the increase in urban female education came from the 12 plus 
group, which now accounts for more than half of all urban women for the most recent cohorts.

@@

Families in urban areas are likely to have fewer children and higher rates of sex 
selection and, therefore, increasingly longer birth intervals over time compared to
rural families.
With the costs of children higher in urban than in rural areas both because of the
lower opportunities for productive work of children and the higher cost of consumption
related to children, fertility is expected to be lower in urban than in rural areas.
Furthermore, access to health care and generally a healthier environment because of,
for example, cleaner water means that urban parents can have fewer children and invest
more in each than in rural areas.%
\footnote{
However, holding parental education constant, there is evidence that children in
Indian slums do worse than they would have in rural areas \citep{Portner2018a}.
}
However, the better access to health care also means that there is easier access to
sex selection and the generally higher income of urban households makes it more
affordable.
[differences in birth intervals without sex selection?]

Wages generally increase with education suggesting that higher female education
should shorten birth spacing and lower fertility, but also increase the use of sex 
selection.
However, female education undoubtedly represents more than the value of women's time.
Two examples are the productions of child health and offspring human capital.
[something on norms?]

First, children of more educated women tend to have better health outcomes and lower 
mortality than less educated women, either from higher income or better health knowledge 
\citep{Rosenzweig1982a,Kovsted2002,Whitworth2002,Maitra2008}.
Lower expected mortality means that fewer births are required to reach a desired
number of surviving children.
Similarly, although short birth spacing generally leads to worse health outcomes, 
the more educated the mother, the more muted this relationship is, which, at least partly,
removes the health outcome incentive for longer birth spacing
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

Second, better-educated women are more efficient at producing child human capital 
\citep{Behrman1999}.
[need something on son preference? Maybe simply use increasing returns to male education]
Hence, families that value sons' education more will tend to both demand better-educated
women as mothers and have fewer children because both allow for higher quality boys.
With the increasing return to male education that India has experienced, this imply
an increased demand for better-educated women combined with lower fertility, and an 
increased use of sex selection when available.
In this case, the associations between higher female education, lower fertility, and more
sex selection are not a causal effect driven by higher wages but instead the result of the 
increased return to male education.

Even as the level of female education has increased, the female labor force 
participation in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
India's female labor force participation is now one of the lowest in the world  
and does not yet show any signs of increasing \citep{Klasen2015,Chatterjee2018}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with the highest percentage working being women 
with either no education or 12 or more years and the lowest being women with 8--11 years 
of education.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

>>>>

This section provides a conceptual framework for understanding how birth spacing changes
and guiding the subsequent analysis.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I argue that female education and area of residence are the most important factors 
to consider in the empirical analysis and discuss how they tie into the three theories.

% Economic conditions
The dramatic improvements in economic conditions in India are best illustrated by the 
doubling of real wages between 1987 and 2011 \citep{Klasen2015}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as a good becomes more expensive, people substitute 
away from that good and towards other goods.
With higher wages, this means that people work more and spend less time on non-wage 
earning activities, such as children or leisure.
The income effect captures that, holding hours worked constant, higher wages increase the 
available income. 
This positive income effect leads people to spend more time on time-intensive activities 
such as children and leisure and less time working.

Empirically, higher female wages generally lowers fertility, whereas higher male 
wages increases fertility \citet{Hotz1997,schultz97}.
This pattern is consistent with women spending substantially more time on child care 
than men, leading to the substitution effect dominating for women whereas the income
effect dominates for men.

In addition to lowering fertility, higher female wages will tend to shorten birth spacing 
if having children requires the mother to reduce her economic activities.
Specifically, shortening birth spacing allows parents to take advantage of economies of 
scale in childrearing---for example, looking after two children simultaneously requires 
less than double the time needed to look after one child \citep{Vijverberg1982,Hotz1997}.
The reduction in the length of birth intervals from higher female wages will be more 
pronounced the more women participate in the labor force rather than working at home as 
economic activities at home are more compatible with child rearing.

% Child outcomes
Parents care not only about how many children they have but also about the outcomes
for the children, and, everything else equal, more children means fewer resources to 
invest per child.
Hence, increasing returns to education provides parents with an incentive to reduce 
fertility and invest more in each of the children's education \citep{Rosenzweig1982a}.
Furthermore, lower expected mortality means that fewer births are required to reach a 
desired number of surviving children and allow parents to invest more in each child.

Longer spacing between births is generally expected to have positive effects on 
child outcomes and if fertility falls because of increased desired for better child
outcomes we would also expect spacing to increase.
The clearest example of birth spacing affecting child outcomes is for health,
where very short spacing---approximately 24 months or less---leads to worse health 
and mortality outcomes for children 
\citep{Whitworth2002,Conde-Agudelo2006,Conde-Agudelo2012}.
Whether longer spacing results in increased parental attention per child and better 
human capital outcome is unclear as the evidence is mixed for developed countries and 
nonexistent for developing countries
\citep{Zajonc1975,Zajonc1976,Razin1980,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Empirically, son preference has only a minor effect on average fertility, but does impact 
the sex composition and family size within families through differential stopping behavior 
\citet{repetto72,leung94,clark00,Basu2010,Barcellos2014}. 
Furthermore, as mentioned above, stronger son preference leads to shorter spacing and  
worse health outcomes for daughters when there are no sons in the household 
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.

@@

Holding son preference constant, lower desired fertility increases the use of sex 
selection \citep{Portner2015b,Jayachandran2017}.  
Furthermore, the higher the parity the more likely the use of sex selection if the 
desired number of sons has not been reached.

% Factors/predictions
Female education and area of residence are the two principal explanatory variables I use
in the empirical analyses.
The main reason is that female education and area of residence are the closest proxies in 
the data for wages and access to and cost of sex selection, none of which are available.
In addition, earlier research show that both higher female education and urbanization are 
associated with lower fertility and increased use of sex selection 
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Finally, India has seen tremendous increases in both education and urbanization.
The proportion of the population living in urban areas almost doubled from 18\%
in 1960 to 35\% in 2019 \citep{United-Nations2019}.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 20 years or older, whether married or not, in the 
four rounds of the NFHS.
Female education has increased substantially in both rural and urban areas,
but is still considerably higher in urban than in rural areas.

@@

Families in urban areas are likely to have fewer children and higher rates of sex 
selection 
Two reasons fertility is lower in urban than rural areas are the higher cost of 
children and the lower mortality risk that comes from better access to health care.%
\footnote{
However, holding parental education constant, there is evidence that children in
Indian slums do worse than they would have in rural areas \citep{Portner2018a}.
}
However, the better access to health care also means that there is easier access to
sex selection and the higher income of urban households makes it more affordable.

Wages generally increase with education suggesting that higher female education
should shorten birth spacing and lower fertility, but also increase the use of sex 
selection.

However, female education undoubtedly represents more than the value of women's time.
Beside propagation of low fertility norms, two examples are the productions of child 
health and offspring human capital.

First, children of more educated women tend to have better health outcomes and lower 
mortality than less educated women, either from higher income or better health knowledge 
\citep{Rosenzweig1982a,Kovsted2002,Whitworth2002,Maitra2008}.
Especially of interest here is that, although short birth spacing generally leads to worse 
health outcomes, the more educated the mother, the more muted this relationship is, which, 
at least partly, removes the health outcome incentive for longer birth spacing
\citep{Molitoris2019}.

Second, better-educated women are more efficient at producing child human capital 
\citep{Behrman1999}.
[need something on son preference? Maybe simply use increasing returns to male education]
Hence, families that value sons' education more will tend to both demand better-educated
women as mothers and have fewer children because both allow for higher quality boys.
With the increasing return to male education that India has experienced, this imply
an increased demand for better-educated women combined with lower fertility, and an 
increased use of sex selection when available.
In this case, the associations between higher female education, lower fertility, and more
sex selection are not a causal effect driven by higher female wages but instead the result 
of the increased return to male education.

Even as the level of female education has increased, the female labor force participation 
in both urban and rural areas has stagnated or decreased and is now one of 
the lowest in the world
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with women with either no education or 12 or more 
years the most likely to participate in the labor force and women with 8--11 years of 
education the least.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}
